What We Eval in the Shadows

### 2021-04-17 paper/main.tex
ck's proofreading of Section 6 and conclusion

The ability to capture current call as a language object (\c{match.call()}),
extract the unevaluated expression from function arguments (\c{substitute()})
and \eval with first class environment makes the R toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \c{data.table}
package defines a DSL for data frame manipulation using the subset operator
\c{[}. This allows one to write compact queries such as \c{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \c{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

@@

\Eval in R is used chiefly for meta-programming purposes and to access
environments other than the local one. In Javascript, patterns are ad-hoc, such
as json loading and parsing, and can often be rewritten without \eval, such as
function or method call, or assignment to a local variable or an object
property. For R, the data suggests that in some cases, the use of \eval in the
\c{X, V, \$, <-, FUN} and the function calls related to code generation could
replaced by more specific functions. For instance, variable lookup in any
environment can be performed with \c{get}, and assignment in any environment,
with \c{assign}. Building a call and executing can be done with \c{do.call}.
Nevertheless, many of the observed \eval usage in R, particularly, the ones
related to statistical modeling and DSLs are more sophisticated than those
reported for JavaScript.

Finally, let us mention an anecdotal use of \eval as a means to bypass some
checks performed by the \c{R CMD CHECK} tool. This tool performs a number of
well-formedness rules before accepting a package to the CRAN repository. Many
checks are performed statically by looking at syntactic constructs in the R
code. Using \eval allows one to bypass such checks and thus submit a package to
CRAN that would otherwise be refused. For example,  it checks that packages
do not call \c{unlockBinding} function.\footnote{It allows one to open and
modify loaded package namespaces.} Packages such as \c{data.table} circumvent
such restrictions by using \eval as the check is only performed statically.

@@

In this paper, we provide a large-scale study of the use of the \eval function
in the R programming language. It is based primarily on a corpus of
\CranRunnableScripts scripts extracted from \CranPackages R packages. As
expected, \eval is used mainly in libraries. We have found just a few \eval
call sites in scripts coming from a large corpus of Kaggle representing code
written by R practitioners. We used an instrumented version of the R virtual
machine, \rdyntrace, to capture all \eval calls traces, including the relevant
side effects that happen within these calls. Several insights can be learned
from the data.
%
First, \eval is widely used in the R core packages that come bundled with the
language. Because of that, there is hardly any R code that would not use it.
\eval is used to implement some basic functionality such as package loading.
Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
the available \CranPackages packages.
%
The code executed by R \eval has unusual expressive power. It can essentially
modify anything anywhere, which is wary for both R developers and R compiler.
The shapes of expressions passed to \eval are extremely diverse, ranging from
simple variable lookups to complex blocks or the creation of new functions.
Luckily, most sites only see a single shape. \eval unlashes its power and shows
up its singularity compared to \eval in other languages with the diversity of
environments in which it can evaluate its expressions. While most of the time,
the \eval environment is the local environment, it is also often the global
environment, a synthetic environment, or even a package environment. Luckily, as
for the expressions passed to \eval, the environment passed to \eval for one
given site is easily predictable with the first call to the site.

>>>>

The ability to capture the current call as a language object (\c{match.call()}),
extract the unevaluated expression from function arguments (\c{substitute()})
and \eval with first class environment makes the R toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \c{data.table}
package defines a DSL for data frame manipulation using the subset operator
\c{[}. This allows one to write compact queries such as \c{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \c{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

@@

\Eval in R is used chiefly for meta-programming purposes and to access
environments other than the local one. In Javascript, patterns are ad-hoc, such
as json loading and parsing, and can often be rewritten without \eval, such as a 
function or method call, or an assignment to a local variable or an object
property. For R, the data suggests that in some cases, the use of \eval in the
\c{X, V, \$, <-, FUN} and the function calls related to code generation could
replaced by more specific functions. For instance, variable lookup in any
environment can be performed with \c{get}, and assignment in any environment,
with \c{assign}. Building a call and executing can be done with \c{do.call}.
Nevertheless, many of the observed \eval usage in R, particularly, the ones
related to statistical modeling and DSLs are more sophisticated than those
reported for JavaScript.

Finally, let us mention an anecdotal use of \eval as a means to bypass some
checks performed by the \c{R CMD CHECK} tool. This tool performs a number of
well-formedness rules before accepting a package to the CRAN repository. Many
checks are performed statically by looking at syntactic constructs in the R
code. Using \eval allows one to bypass such checks and thus submit a package to
CRAN that would otherwise be refused. For example,  it checks that packages
do not call the \c{unlockBinding} function.\footnote{It allows one to open and
modify loaded package namespaces.} Packages such as \c{data.table} circumvent
such restrictions by using \eval as the check is only performed statically.

@@

In this paper, we provide a large-scale study of the use of the \eval function
in the R programming language. It is based primarily on a corpus of
\CranRunnableScripts scripts extracted from \CranPackages R packages. As
expected, \eval is used mainly in libraries. We have found just a few \eval
call sites in scripts coming from a large corpus of Kaggle representing code
written by R practitioners. We used an instrumented version of the R virtual
machine, \rdyntrace, to capture all \eval calls, including the relevant
side effects that happen within these calls. Several insights can be learned
from the data.
%
First, \eval is widely used in the R core packages that come bundled with the
language. Because of that, there is hardly any R code that would not use it.
\eval is used to implement some basic functionality such as package loading.
Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
the available \CranPackages packages.
%
The code executed by R \eval has unusual expressive power. It can essentially
modify anything anywhere, which is wary for both R developers and R compiler.
The shapes of expressions passed to \eval are extremely diverse, ranging from
simple variable lookups to complex blocks or the creation of new functions.
Luckily, most sites only see a single shape. \eval unlashes its power and shows
up its singularity compared to \eval in other languages with the diversity of
environments in which it can evaluate its expressions. While most of the time,
the \eval environment is the local environment, it is also often the global
environment, a synthetic environment, or even a package environment. Luckily, as
for the expressions passed to \eval, the environment passed to \eval for one
given site is easily predictable with the first call to the site.
